---
title: "RepDataAssignment1-PA"
author: "Peter Anger"
date: "4/20/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This markdown file was produced for the first assignment in the
Reproducible Research course.

#####Load the the activity data  
- read in the csv file
```{r load_data, echo = TRUE}
act<-read.csv("activity.csv",colClasses="character")
head(act)
```
- set appropriate classes for ech column of data  
```{r set_classes, echo = TRUE}
act[,1] <- as.numeric(act[,1])
act[,3] <- as.numeric(act[,3])
str(act)
```
  
#####Look at daily values:  
- make a clean copy of the data (removing rows with NA data)
```{r clean_data, echo = TRUE}
cleanAct <- na.omit(act)
```
- compute the total steps per day
```{r total_daily_steps, echo = TRUE}
totalStepsDay <- tapply(cleanAct$steps,cleanAct$date,sum)
head(totalStepsDay)
```
- generate a histogram of total steps per day
```{r hist_daily_steps, echo = FALSE}
hist(totalStepsDay,main="",xlab="Total Steps per Day")
```
- identify the mean and median total steps per day
```{r mean_median_steps, echo = TRUE}
mean(totalStepsDay)
median(totalStepsDay)
```

#####Look at the daily activity pattern:  
- compute the number of steps per interval averaged across days
```{r mean_steps_per_interval, echo = TRUE}
meanStepsInterval <-tapply(cleanAct$steps,cleanAct$interval,mean)
meanStepsInterval<-data.frame(interval=as.numeric(rownames(meanStepsInterval)),steps=meanStepsInterval)
head(meanStepsInterval)
```
- create a time series plot of the steps per interval
```{r plot_interval_steps, echo = FALSE}
plot(meanStepsInterval,type="l", xlab = "Interval", ylab = "Steps")
```
- identify the interval with the highest average step count
```{r which_highest_interval_steps, echo = TRUE}
meanStepsInterval$interval[which.max(meanStepsInterval$steps)]
```
- and the corresponding step count
```{r max_interval_steps, echo = TRUE}
max(meanStepsInterval$steps)
```

#####Fill in mising values:  
- identify the number of rows with NA data
``` {r missing_values, echo = TRUE}
nrow(act)-nrow(cleanAct)
```
- create a copy of the dataset
``` {r copy_data, echo = TRUE}
fixedAct<-act
```
- replace NA values with the mean number of steps for the given
interval (averaged across all days)
``` {r impute_missing_values, echo = TRUE}
for (row in 1:nrow(fixedAct)) {
    if (is.na(fixedAct$steps[row])) {
        fixedAct$steps[row] <- round(meanStepsInterval$steps[(as.character(fixedAct$interval[row]))])
    }
}
```
#####Reanalyze the "fixed" daily data  
- compute total steps per day
```{r total_fixed_daily steps, echo = TRUE}
totalStepsDay2 <- tapply(fixedAct$steps,fixedAct$date,sum)
head(totalStepsDay2)
```
- generate a histogram of daily steps
```{r hist_fixed_daily_steps, echo = FALSE}
hist(totalStepsDay2,main="",xlab="Total Steps per Day")
```
- identify the mean daily steps
```{r mean_fixed_daily_steps, echo = TRUE}
mean(totalStepsDay2)
```
- identify the median daily steps
```{r median_fixed_daily_steps_daily, echo = TRUE}
median(totalStepsDay2)
```
- compare to cleaned data 
```{r fixed_vs_cleaned_mean_median, echo = TRUE}
mean(totalStepsDay2) / mean(totalStepsDay)
median(totalStepsDay2) / median(totalStepsDay)
```

#####Reanalyze the "fixed" interval data  
- compute count of fixed steps per interval averaged across all days
```{r fixed_interval_steps, echo = TRUE}
meanStepsInterval2 <-round(tapply(fixedAct$steps,fixedAct$interval,mean))
meanStepsInterval2<-data.frame(interval=as.numeric(rownames(meanStepsInterval2)),steps=meanStepsInterval2)
head(meanStepsInterval2)
```
- create a time series plot of the steps per interval
```{r plot_fixed_interval_steps, echo = FALSE}
plot(meanStepsInterval2,type="l")
```
- identify the interval with the highest average step count
```{r highest_fixed_interval, echo = TRUE}
meanStepsInterval2$interval[which.max(meanStepsInterval2$steps)]
max(meanStepsInterval2$steps)
```
- compare to cleaned data 
```{r fixed_vs_clean_interval_steps, echo = TRUE}
meanStepsInterval2$interval[which.max(meanStepsInterval2$steps)] / meanStepsInterval$interval[which.max(meanStepsInterval$steps)]
max(meanStepsInterval2$steps) / max(meanStepsInterval$steps)
```

#####Organize data by weekday Vs weekend  
- add a factor indicating whether a given date is a weekday or a weekend
```{r add_daytype_factor, echo = TRUE}
type<-as.character()
days <- c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday")
for (i in 1:5) type[days[i]] <- "weekday"
for (i in 6:7) type[days[i]] <- "weekend"
fixedAct$dayType <- type[weekdays(as.Date(fixedAct$date))]
fixedAct$dayType <- as.factor(fixedAct$dayType)
```
- create subsets of the data for weekdays vs weekends
```{r subset_by_daytype, echo = TRUE}
fixedActWD<-subset(fixedAct,as.character(fixedAct$dayType)=="weekday")
meanStepsIntervalWD <-round(tapply(fixedActWD$steps,fixedActWD$interval,mean))
meanStepsIntervalWD<-data.frame(interval=as.numeric(as.character(rownames(meanStepsIntervalWD))),steps=meanStepsIntervalWD)
fixedActWE<-subset(fixedAct,as.character(fixedAct$dayType)=="weekend")
meanStepsIntervalWE <-round(tapply(fixedActWE$steps,fixedActWE$interval,mean))
meanStepsIntervalWE<-data.frame(interval=as.numeric(as.character(rownames(meanStepsIntervalWE))),steps=meanStepsIntervalWE)
```

#####Compare weekday to weekend steps  
-Generate a time series panel plot per the read.me example that compares the mean steps/interval for weekdays and weekends.
``` {r panelplot, echo = FALSE}
par(mfrow=c(2,1),cex=.75,mai=c(0,0.25,0.25,0),omi=c(0.75,0.5,0,0.2),col="blue",ann=TRUE)
plot(meanStepsIntervalWD$interval,meanStepsIntervalWD$steps,type="l",xaxt="n",ylab="",xlab="",main="WEEKDAY")
plot(meanStepsIntervalWE$interval,meanStepsIntervalWE$steps,type="l",ylab="",xlab="",main="WEEKEND",mtext("Interval",side=1,line=17,col="black",cex=1.25),mtext("Number of Steps",adj=2.2,side=2,line=3,col="black",cex=.75))
```